---
title: "Lab 09: Algorithmic Bias"
author: "Cynthia Deng"
date: "3/2/2026"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```


## Exercise 1

The dataset has `r nrow(compas)` rows and `r ncol(compas)` columns.

## Exercise 2
```{r}
n_distinct(compas$id)                 
n_distinct(compas$name)  
n_distinct(compas$first)  
n_distinct(compas$last)  
```
There are 7,214 rows in the dataset, but only 7,158 unique defendant IDs. The number of rows exceeds the number of unique defendants. There could be 2 possible explanations. 1. Some individuals appear more than once. 2. Some people have the same name.

## Exercise 3
```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_histogram(binwidth = 1, color = "black") +
  labs(title = "Distribution of COMPAS Decile Scores",
       x = "Decile Score",
       y = "Count")
```
```{r}
# Race
ggplot(compas, aes(x = race)) +
  geom_bar() +
  labs(title = "Distribution of Defendants by Race", x = "Race", y = "Count")

# Sex
ggplot(compas, aes(x = sex)) +
  geom_bar() +
  labs(title = "Distribution of Defendants by Sex", x = "Sex", y = "Count")

# Age Category
ggplot(compas, aes(x = age_cat)) +
  geom_bar() +
  scale_x_discrete(limits = c("Less than 25", "25 - 45", "Greater than 45")) +
  labs(title = "Distribution of Defendants by Age Category", x = "Age Category", y = "Count")
```

## Exercise 5
```{r}
ggplot(compas, aes(x = factor(decile_score), y = two_year_recid)) +
  stat_summary(fun = mean, geom = "bar") +
  labs(title = "Recidivism Rate by COMPAS Decile Score",
       x = "Decile Score",
       y = "Proportion Who Reoffended")
```

Yes, higher COMPAS risk scores do correspond to higher rates of recidivism. The chart shows a clear and consistent positive trend

## Exercise 6
```{r}
compas <- compas %>%
  mutate(compas_classification = case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score <= 4 & two_year_recid == 0 ~ "TN",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    TRUE ~ NA_character_  # scores 5-6 are excluded
  )) 
compas %>%
  count(compas_classification) %>%
  mutate(percent = round(n / sum(n) * 100, 1))
```

## Exercise 7
```{r}
compas %>%
  filter(!is.na(compas_classification)) %>%
  summarise(accuracy = mean(compas_classification %in% c("TP", "TN")))
```
Among defendants with a clear low (≤4) or high (≥7) risk score, COMPAS correctly classified 68.4% of cases, with false negatives (16.9%) outnumbering false positives (8.9%), suggesting the algorithm is more likely to underestimate risk than overestimate it.

## Exercise 8
```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  ggplot(aes(x = factor(decile_score), fill = race)) +
  geom_bar(position = "fill") +
  labs(title = "Distribution of COMPAS Risk Scores by Race",
       x = "Decile Score",
       y = "Proportion",
       fill = "Race")
```

The distribution of COMPAS risk scores differs notably between Black and white defendants. At low decile scores (1–2), Caucasian defendants make up a larger share, with white defendants representing about 35–50% of those scores. However, as decile scores increase, the proportion of African-American defendants grows steadily. At scores of 8–10, Black defendants make up roughly 80% of all defendants at those risk levels. 

## Exercise 9
```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race) %>%
  summarise(
    total = n(),
    high_risk = sum(decile_score >= 7),
    percent_high_risk = mean(decile_score >= 7) * 100
  )
```
There is a clear racial disparity in high-risk classifications. 38.6% of Black defendants were classified as high risk (decile score ≥ 7), compared to only 17.1% of white defendants

## Exercise 10
```{r}
# False Positive Rate by race
fpr <- compas %>%
  filter(race %in% c("African-American", "Caucasian"),
         two_year_recid == 0) %>%
  group_by(race) %>%
  summarise(
    false_positive_rate = mean(decile_score >= 7) * 100
  )

# False Negative Rate by race
fnr <- compas %>%
  filter(race %in% c("African-American", "Caucasian"),
         two_year_recid == 1) %>%
  group_by(race) %>%
  summarise(
    false_negative_rate = mean(decile_score <= 4) * 100
  )

fpr
fnr
```
## Exercise 11
```{r}
# Calculate both metrics
metrics <- bind_rows(
  compas %>%
    filter(race %in% c("African-American", "Caucasian"), two_year_recid == 0) %>%
    group_by(race) %>%
    summarise(rate = mean(decile_score >= 7) * 100, metric = "False Positive Rate"),
  compas %>%
    filter(race %in% c("African-American", "Caucasian"), two_year_recid == 1) %>%
    group_by(race) %>%
    summarise(rate = mean(decile_score <= 4) * 100, metric = "False Negative Rate")
)

# Visualize
ggplot(metrics, aes(x = metric, y = rate, fill = race)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "False Positive and False Negative Rates by Race",
       x = "Metric", y = "Rate (%)", fill = "Race")
```

Black defendants had a false positive rate of ~25%, more than double the ~9% for white defendants, while white defendants had a much higher false negative rate (~48%) compared to Black defendants (~28%). 

## Exercise 12
```{r}
ggplot(compas, aes(x = decile_score, y = priors_count, color = race)) +
  geom_jitter(alpha = 0.05, width = 0.2, height = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Risk Score vs Prior Convictions by Race",
    x = "COMPAS Decile Score",
    y = "Number of Prior Convictions",
    color = "Race"
  ) +
  theme_minimal()
```

## Exercise 13
```{r}
compas %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, decile_score) %>%
  summarise(recidivism_rate = mean(two_year_recid)) %>%
  ggplot(aes(x = decile_score, y = recidivism_rate, color = race)) +
  geom_line() +
  geom_point() +
  labs(title = "Calibration: Recidivism Rate by Decile Score and Race",
       x = "Decile Score",
       y = "Proportion Who Reoffended",
       color = "Race") +
  theme_minimal()
```

The lines track closely together across all decile scores, meaning a given score predicts similar recidivism rates for both races.

## Exercise 14, 15, 16
To address the disparities observed, I would reconsider which variables are included in the model. Features like neighborhood or employment status may introduce bias by correlating with race due to historical inequality. Additionally, I would explore using separate, group-specific models to ensure predictions are equally accurate across demographic groups, and establish clear guidelines on how scores should and should not be used by judges to prevent over-reliance on algorithmic outputs.

Designing a fair algorithm inevitably involves trade-offs, as prioritizing one type of fairness often comes at the cost of another. For example, reducing false positives for one group may increase false negatives for another, and these errors carry different ethical costs — incorrectly labeling someone as high risk has different consequences than failing to identify someone who reoffends.

Balancing competing fairness definitions requires difficult value judgments. For example, whether it is worse to incorrectly flag an innocent person as high risk or to miss someone who will reoffend. These are not purely technical choices but reflect deeper ethical priorities. Additionally, implementing fairness improvements requires resources, which raises questions about how to weigh fairness goals against practical constraints. Ultimately, what counts as "fair" is shaped by the values of the legal system and society applying it, meaning there may be no universal answer, the answer may be dependent on the culture and context.
